---
title: "Getting started with rk8s"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with rk8s}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

rk8s is an R client for the Kubernetes API. It has two layers:

* **A high-level verb API** — `kube_get()`, `kube_apply()`, `kube_logs()`,
  `kube_watch()`, … This is what most users want. You never have to type
  `V1Deployment$new(...)` to do everyday work.
* **A low-level typed/generated API** — `CoreV1Api`, `AppsV1Api`,
  `V1Pod`, `V1Deployment`, … 944 operations and 635 model classes
  generated from the Kubernetes OpenAPI spec, mirroring the official
  Python client and Go client-go. There when you need it.

A `Watch` helper, `Informer`/`Lister` for client-go-style controllers,
`DynamicClient` for CustomResources, and `run_as_leader()` for leader
election sit alongside both layers.

## Installing

```r
# remotes::install_github("ian-flores/rk8s")
install.packages("rk8s_0.2.0.tar.gz", repos = NULL)
```

## Authenticating

```{r}
library(rk8s)
k <- kube()                                   # ~/.kube/config
# k <- kube(context = "my-dev")
# k <- kube(config_file = "~/configs/prod.yaml")
# k <- kube(namespace = "my-app")             # change default namespace
```

Inside a pod running as a ServiceAccount:

```{r}
k <- kube(client = ApiClient$new(load_incluster_config()))
```

`load_kube_config()` understands everything `kubectl` does: bearer
tokens, token files, basic auth, inline base64 client certs, on-disk
PEM files, and `exec` credential plugins (EKS's `aws-iam-authenticator`,
GKE's `gke-gcloud-auth-plugin`, any OIDC helper).

## The verb API

Resource references can be:

* a kind alias: `"pod"`, `"pods"`, `"po"`, `"deploy"`, `"svc"`, `"ing"`,
  `"crd"`, …
* `"kind/name"`: `"deployment/nginx"`
* explicit `"apiVersion:Kind"`: `"apps/v1:Deployment"`
* a `c(apiVersion, Kind)` vector: `c("apps/v1", "Deployment")`

### Get / list

```{r}
kube_get(k, "pods", namespace = "default")              # list
kube_get(k, "pod", "nginx-abc")                          # single
kube_get(k, "deployment/nginx")                          # alt syntax
kube_get(k, "pods", label_selector = "app=nginx")
kube_get(k, "apps/v1:Deployment", "nginx")               # explicit GVK
```

`kube_get()` returns an `rk8s_object` (single) or `rk8s_list` (collection).
A built-in `as.data.frame()` method gives you a tabular view:

```{r}
pods <- kube_get(k, "pods")
head(as.data.frame(pods))
#>     namespace                    name kind     age   status
#> 1     default                  nginx-1  Pod    12m  Running
#> 2     default                  nginx-2  Pod    12m  Running
#> 3 kube-system          coredns-abc-xyz  Pod      4h  Running
```

### Apply (server-side apply)

`kube_apply()` accepts a path, a YAML/JSON string, an R `list`, or a
generated model. Multi-document YAML works — each document is routed
to the correct endpoint via discovery:

```{r}
# From a file
kube_apply(k, "deployment.yaml")

# From an R list (the most natural form for programmatic generation)
kube_apply(k, list(
  apiVersion = "v1", kind = "ConfigMap",
  metadata = list(name = "demo"),
  data = list(hello = "world")
))

# From a multi-doc YAML string
kube_apply(k, "
apiVersion: v1
kind: ConfigMap
metadata: {name: demo-cm}
data: {greeting: hello}
---
apiVersion: v1
kind: Service
metadata: {name: demo-svc}
spec:
  selector: {app: demo}
  ports: [{port: 80, targetPort: 8080}]
")

# Force-take ownership of conflicting fields (server-side apply force):
kube_apply(k, "deployment.yaml", force = TRUE)
```

Field manager defaults to `"rk8s"`; override with `field_manager =`.

### Delete, scale, logs, watch

```{r}
kube_delete(k, "pod", "nginx-abc")
kube_delete(k, "deployment/nginx")
kube_delete(k, "deployment/nginx", grace_period = 0)

kube_scale(k, "deployment", "nginx", replicas = 5)
kube_scale(k, "statefulset/postgres", replicas = 3)

cat(kube_logs(k, "nginx-abc", tail = 50))
kube_logs(k, "nginx-abc", container = "sidecar", follow = TRUE)

kube_watch(k, "pods",
  on_event = function(type, obj) {
    message(type, " ", obj$metadata$name)
    TRUE  # keep watching; return FALSE to stop
  },
  timeout_seconds = 30
)
```

### Anything custom

`kube_apply()` works for CRs out of the box (any `apiVersion+kind`),
and `kube_get()` resolves them via the dynamic client. So a typical
controller is:

```{r}
foos <- kube_list(k, "example.com/v1:Foo", namespace = "default")
for (f in foos$items) {
  if (is.null(f$status$ready) || !isTRUE(f$status$ready)) {
    kube_apply(k, list(
      apiVersion = "example.com/v1", kind = "Foo",
      metadata = list(name = f$metadata$name),
      status = list(ready = TRUE)
    ))
  }
}
```

## Watching with a local cache: Informer + Lister

When you want to react to changes without re-listing the API server
every tick, use `Informer`. One watch feeds an in-memory cache plus
your event handlers; `Lister` gives you O(1) lookups.

```{r}
inf <- Informer$new(
  k$client,
  resource_path = "/api/v1/namespaces/default/pods",
  object_from_list = V1Pod$from_list
)
inf$add_event_handler(
  on_add    = function(obj) message("add: ", obj$metadata$name),
  on_update = function(old, new) {
    if (!identical(old$status$phase, new$status$phase))
      message(new$metadata$name, ": ",
              old$status$phase, " -> ", new$status$phase)
  },
  on_delete = function(obj) message("gone: ", obj$metadata$name)
)
inf$run(stop_seconds = 120)

pods <- inf$lister()
pods$get("nginx")    # cache lookup, no API call
pods$list()
```

## Leader election

For controllers that need exactly-one-active across replicas:

```{r}
run_as_leader(
  k$client,
  lease_name = "my-controller",
  lease_namespace = "kube-system",
  identity = paste0(Sys.info()[["nodename"]], "-", Sys.getpid()),
  on_started_leading = function() { message("leading"); controller_loop() },
  on_stopped_leading = function() message("stopped leading")
)
```

A `callr`-driven background renewer keeps the lease fresh; on failure
or `Ctrl-C`, `on_stopped_leading` fires.

## Errors

API failures surface as `rk8s_api_error` conditions whose `$exception`
is an `ApiException` with the HTTP status, reason, headers, and parsed
`metav1.Status`.

```{r}
pod <- tryCatch(
  kube_get(k, "pod", "nope"),
  rk8s_api_error = function(e) {
    if (e$exception$status == 404L) NULL else stop(e)
  }
)
```

## Dropping to the typed/generated layer

Any time `kube_*` doesn't expose what you need, drop down. The verb
client carries the underlying `ApiClient` and `DynamicClient`:

```{r}
# Typed:
core <- CoreV1Api$new(k$client)
pods <- V1PodList$from_list(core$list_pod_for_all_namespaces())

# Or build a typed model explicitly:
dep <- V1Deployment$new(
  api_version = "apps/v1", kind = "Deployment",
  metadata = V1ObjectMeta$new(name = "nginx"),
  spec = V1DeploymentSpec$new(
    replicas = 3L,
    selector = V1LabelSelector$new(match_labels = list(app = "nginx")),
    template = V1PodTemplateSpec$new(
      metadata = V1ObjectMeta$new(labels = list(app = "nginx")),
      spec = V1PodSpec$new(containers = list(
        V1Container$new(name = "nginx", image = "nginx:1.25"))))))
AppsV1Api$new(k$client)$create_namespaced_deployment(
  namespace = "default", body = dep)
```

## Regenerating from a newer Kubernetes spec

```bash
curl -L -o tools/gen/spec/swagger.json \
  https://raw.githubusercontent.com/kubernetes/kubernetes/release-1.31/api/openapi-spec/swagger.json
Rscript tools/gen/generate.R tools/gen/spec/swagger.json R
R CMD build .
```

The generator (`tools/gen/`) reads OpenAPI v2 JSON and emits R6 model
classes plus per-tag API classes — about 700 R files. NAMESPACE entries
between sentinel lines are managed automatically.
