---
title: "Getting started with rk8s"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with rk8s}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

rk8s is an R client for the Kubernetes API. It mirrors the architecture of
the official Python client and Go client-go: a small hand-written runtime
(configuration, HTTP client, watch, dynamic client, informer, leader
election, utilities) plus typed API and model classes code-generated
from the Kubernetes OpenAPI specification.

This vignette walks through the essentials. Every code chunk assumes a
reachable Kubernetes API server and a valid kubeconfig (or in-cluster
service account).

## Installing

```r
# From a local tarball:
install.packages("rk8s_0.1.0.tar.gz", repos = NULL)

# Or from GitHub:
# remotes::install_github("ian-flores/rk8s")
```

## Authenticating

Most users rely on their existing kubeconfig:

```{r}
library(rk8s)

client <- new_client_from_config()            # ~/.kube/config
# client <- new_client_from_config(context = "my-dev")
# client <- new_client_from_config("~/configs/prod.yaml")
```

Inside a pod running as a ServiceAccount, use `load_incluster_config()`:

```{r}
client <- ApiClient$new(load_incluster_config())
```

`load_kube_config()` supports everything `kubectl` does: bearer tokens,
token files, basic auth, inline base64 client certs, on-disk PEM files,
and `exec` credential plugins (EKS's `aws-iam-authenticator`, GKE's
`gke-gcloud-auth-plugin`, any OIDC helper).

## Typed API classes

The generated API classes give you CRUD + subresource methods for every
resource:

```{r}
core <- CoreV1Api$new(client)

# List pods across all namespaces
pods <- V1PodList$from_list(core$list_pod_for_all_namespaces())
for (p in pods$items) {
  cat(p$metadata$namespace, "/", p$metadata$name, "\n", sep = "")
}

# Read a specific pod
pod <- core$read_namespaced_pod(namespace = "default", name = "nginx")

# Delete
core$delete_namespaced_pod(namespace = "default", name = "nginx")
```

### Creating resources

Build resources with the typed model classes. They round-trip to and
from JSON cleanly:

```{r}
apps <- AppsV1Api$new(client)
dep <- V1Deployment$new(
  api_version = "apps/v1", kind = "Deployment",
  metadata = V1ObjectMeta$new(name = "nginx"),
  spec = V1DeploymentSpec$new(
    replicas = 3L,
    selector = V1LabelSelector$new(match_labels = list(app = "nginx")),
    strategy = V1DeploymentStrategy$new(
      type = "RollingUpdate",
      rolling_update = V1RollingUpdateDeployment$new(
        max_surge = IntOrString$new(value = "25%"),
        max_unavailable = IntOrString$new(value = 0L))),
    template = V1PodTemplateSpec$new(
      metadata = V1ObjectMeta$new(labels = list(app = "nginx")),
      spec = V1PodSpec$new(containers = list(
        V1Container$new(name = "nginx", image = "nginx:1.25",
                         ports = list(V1ContainerPort$new(container_port = 80L))))))))

apps$create_namespaced_deployment(namespace = "default", body = dep)
```

You can also pass a plain nested `list` as `body` — model classes are
convenience, not requirement.

## Watching

`Watch` streams events from any list endpoint. The callback receives
`type` (`ADDED`/`MODIFIED`/`DELETED`/`BOOKMARK`) and the object. Return
`FALSE` to stop; the stream automatically reconnects with the last
observed `resourceVersion`, re-listing on 410 Gone.

```{r}
w <- Watch$new()
w$stream(
  client,
  resource_path = "/api/v1/namespaces/default/pods",
  callback = function(type, object) {
    cat(type, object$metadata$name, "\n")
    TRUE
  },
  timeout_seconds = 60
)
```

## Informer + Lister

When you want to react to changes *without* re-listing the API server
every tick, use `Informer`: a local cache kept in sync by one watch,
plus event handlers. This is how controllers are built.

```{r}
inf <- Informer$new(
  client,
  resource_path = "/api/v1/namespaces/default/pods",
  object_from_list = V1Pod$from_list
)
inf$add_event_handler(
  on_add    = function(obj) message("add: ", obj$metadata$name),
  on_update = function(old, new) {
    if (!identical(old$status$phase, new$status$phase))
      message(new$metadata$name, ": ", old$status$phase, " -> ", new$status$phase)
  },
  on_delete = function(obj) message("gone: ", obj$metadata$name)
)
inf$run(stop_seconds = 120)

pods <- inf$lister()
pods$get("nginx")       # O(1) cache lookup, no API call
pods$list()             # current cache snapshot
```

## Dynamic client (CustomResources and anything untyped)

For CRDs (or when you'd rather not import a typed class), use
`DynamicClient`. It discovers the REST path for any `apiVersion+kind`:

```{r}
dc <- DynamicClient$new(client)

# Built-in:
pods <- dc$resource("v1", "Pod")
pods$list(namespace = "kube-system")
pods$get("kube-dns-abc", namespace = "kube-system")

# CustomResource:
foos <- dc$resource("example.com/v1", "Foo")
foos$list(namespace = "default")
foos$create(list(
  apiVersion = "example.com/v1", kind = "Foo",
  metadata = list(name = "my-foo"),
  spec = list(bar = 1)
), namespace = "default")
```

## Apply a YAML manifest

`create_from_yaml()` routes each document to the right endpoint via the
dynamic client, like `kubectl apply -f`:

```{r}
create_from_yaml(client, yaml_file = "manifest.yaml")
```

## Leader election

For controllers that need exactly-one-active semantics across replicas:

```{r}
run_as_leader(
  client,
  lease_name = "my-controller",
  lease_namespace = "kube-system",
  identity = paste0(Sys.info()[["nodename"]], "-", Sys.getpid()),
  on_started_leading = function() { message("leading"); controller_loop() },
  on_stopped_leading = function() message("stopped leading")
)
```

A background renewer (via the optional `callr` package) keeps the lease
fresh while your controller loop runs; on failure or `Ctrl-C`, the
`on_stopped_leading` callback fires.

## Handling errors

API failures surface as `rk8s_api_error` conditions whose `$exception`
is an `ApiException` with the HTTP status, reason phrase, headers, and
parsed `metav1.Status`.

```{r}
pod <- tryCatch(
  core$read_namespaced_pod(namespace = "default", name = "nope"),
  rk8s_api_error = function(e) {
    if (e$exception$status == 404L) NULL else stop(e)
  }
)
```

## Regenerating from a newer OpenAPI spec

The generator in `tools/gen/` produces the typed classes (`gen_model_*.R`
and `gen_api_*.R`). To update to a newer Kubernetes version:

```bash
curl -L -o tools/gen/spec/swagger.json \
  https://raw.githubusercontent.com/kubernetes/kubernetes/release-1.31/api/openapi-spec/swagger.json
Rscript tools/gen/generate.R tools/gen/spec/swagger.json R
R CMD build .
```

See `tools/gen/README.md` for generator internals.
